\documentclass[problem]{mcs}

\begin{pcomments}
  \pcomment{PS_MIT_Harvard_degree_value_monthly}
  \pcomment{subsumes PS_MIT_Harvard_degree_value}
  \pcomment{ARM 4/28/18}
\end{pcomments}

\pkeywords{
  geometric_sum
  series
  summation
  interest
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem starts here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{problem}
Is a Harvard degree really worth more than an MIT degree?  Let us say
that a new Harvard graduate starts with a monthly salary of \$10K and
gets a \$1K raise every month.  So after one month, the Harvard grad
gets paid \$10K, at the end of the second month they get paid \$11K, at
the end of the third month \$12K, \dots.

On the other hand, a new MIT graduate starts with a monthly salary of
\$6,000, but gets a 2\% raise every month.  So after one month, the
MIT grad gets paid \$6K, at the end of the second month they get paid
$1.02\cdot \$6K = \$6,120$, at the end of the third month $1.02\cdot
\$6120 = \$6242.40$, \dots.  Assume inflation is a fixed $0.3\%$ per
month.  That is, \$1.003 a month from now is worth \$1.00 today.

\begin{problemparts}

\ppart Write a closed form formula for the total number of thousands
of dollars a Harvard grad will earn during the $n$ years after
graduation.  What is the total when $n=25$?

\begin{solution}
\[
\mbox{Hvd-total}_n  =\sum_{1}^{12n} 10 + (i-1) = 120n + 12n(12(n-1))/2 = 72n^2 + 114n.
\]
For $n = 25$, the total is \$47850K.
\end{solution}

\problempart Write a closed form formula for the value in today's
dollars of a Harvard degree, assuming the Harvard grad will work for
$n$ years following graduation.  What is the value when $n=25$?

\begin{solution}
One dollar at the end of month $i$ is worth $r^i$ in today's currency,
where
\[
r \eqdef \frac{1}{1.003}
\]
So
\[
\mbox{Hvd-val}_n  = \sum_{1}^{12n} (10+ (i-1))r^i
     = 10r\sum_{i=0}^{12n-1} r^i + r\sum_{i=0}^{12n-1} i r^i.
\]
Using the formula\inbook{~(\bref{geometric-sum-n})} for geometric
series, we have
\[
\sum_{i=0}^{12n-1} r^i = \frac{(1- r^{12n})}{1 - r}.
\]
Also, applying the formula\inbook{~(\bref{sumixi})}\inhandout{ from the
  text} for a sum of terms of the form $ir^i$, we have
\[
\sum_{i=0}^{12n-1} ir^i = \frac{{r - 12n r^{12n} + (12n-1) r^{12n+1}}}{(1-r)^2},
\]
so
\[
\mbox{Hvd-val}_n
= 10r \frac{(1- r^{12n})}{1 - r} +
    r \frac{{r - 12n r^{12n} + (12n-1) r^{12n+1}}}{(1-r)^2}.
\]

\begin{staffnotes}
\begin{verbatim}
(define r (/ 1 1.003))

(define (h-val n)
  (+ (* 10 r (/ (- 1 (expt r (* 12 n))) (- 1 r)))
     (* r (/ (+ r
                (* (- (* 12 n))   (expt r (* 12 n)))
                (* (- (* 12 n) 1) (expt r (+ (* 12 n) 1))))
             (* (- 1 r) (- 1 r))))))
(h-val 25)
;Value: 27140.293424774896
\end{verbatim}
\end{staffnotes}

For $n = 25$, the value is \$27140K, which is about half the total
paid out over 25 years.

\end{solution}

\problempart Likewise for an MIT degree.

\begin{solution}
\[
\mbox{MIT-tot}_n = 6 \sum_{i=1}^{12n }1.02^i = 6\paren{\paren{\sum_{i=0}^{12n} 1.02^i}- 1}
= 6\paren{\frac{1 - 1.02^{12n+1}}{1 - 1.02} - 1}.
\]

\begin{staffnotes}
\begin{verbatim}
(define p 1.02)

(define (m-tot n)
    (* 6
       (- (/ ( - 1 (expt p (+ (* 12 n) 1)))
             (- 1 p))
          1)))
(m-tot 25)
;Value: 116045.75946682901
\end{verbatim}
\end{staffnotes}
For $n = 25$, the MIT total is \$116046K, which is almost two and a
half times the Harvard total.


\[
\mbox{MIT-val}_n = 6\sum_{i=1}^{12n} {1.02}^i r^i
= 6\paren{\paren{\sum_{i=0}^{12n} (1.02r)^i}- 1}
= 6\paren{\frac{1 - (1.02r)^{12n+1}}{1 - 1.02r} - 1}.
\]

\begin{staffnotes}
\begin{verbatim}
(define p 1.02)

(define r (/ 1 1.003))

(define (m-val n)
    (* 6
       (- (/ ( - 1 (expt (* p r) (+ (* 12 n) 1)))
             (- 1 (* p r)))
          1)))
(m-val 25)
;Value: 55368.085473375395
\end{verbatim}
\end{staffnotes}

For $n = 25$, the value the MIT degree is \$55368K, which is a little
more than twice as valuable as a Harvard degree! (But we took that for
granted, right?)
\end{solution}

\end{problemparts}
\end{problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Problem ends here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
